---
title: "HAR Predict Activity"
author: "Ash Chakraborty"
date: "December 21, 2015"
output: html_document
---  

```{r global ECHO=FALSE, message=FALSE, warning=FALSE, error=FALSE}
library(caret)
library(ggplot2)
library(GGally)
library(foreach)
library(doParallel)
train.url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test.url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```  

# OVERVIEW  

"Human Activity Recognition" (HAR) exercise. Use data collected from a wearable
device to train a couple of machine learning models so that they can "learn" to
predict the "type" of activity ('classe') given a test data set.  

The link and the project are here: http://groupware.les.inf.puc-rio.br/har

# DATA PREVIEW  

Training data is at: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv  

We'll need a cross validation set, since the testing set is provided separately, here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

So, let's read in data, and create a Cross Validation set:  

```{r}
raw <- read.csv(train.url, header = T, na.strings = c(""," ","#DIV/0!","NA"))
trainInds <- createDataPartition(raw$classe, p=.7, list = F)
train <- raw[trainInds, -c(1:4)] #names, ids, timestamps aren't good predictors
cv <- raw[-trainInds, -c(1:4)]

```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
